---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# phutil

<!-- badges: start -->
[![R-CMD-check](https://github.com/tdaverse/phutil/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/tdaverse/phutil/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/tdaverse/phutil/graph/badge.svg)](https://app.codecov.io/gh/tdaverse/phutil)
<!-- badges: end -->

The goal of the **phutil** package is to provide utility functions for
persistence data analysis. In particular, the package defines a data structure
for hosting persistence data. The package is part of the
[*tdaverse*](https://github.com/tdaverse) suite of packages, which aims to
provide a consistent interface for topological data analysis in R that nicely
integrates with the [*tidymodels*](https://www.tidymodels.org) ecosystem.

## Installation

You can install the development version of **phutil** from
[GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("tdaverse/phutil")
```

## Example

The package currently provides a new data structure for hosting persistence
data. The data set `noisy_circle_points` is a simulated data set consisting of
100 points sampled from a circle with additive Gaussian noise using a standard
deviation of $0.05$:

```{r}
library(phutil)
head(noisy_circle_points)
```

The data set `noisy_circle_ripserr` is a persistence diagram computed using the
[**ripserr**](https://tdaverse.github.io/ripserr/) package with the
[`ripserr::vietoris_rips()`](https://tdaverse.github.io/ripserr/reference/vietoris_rips.html)
and stored as an object of class 'PHom', which is a light wrapper around a data
frame with 3 variables:

```{r}
head(noisy_circle_ripserr)
```

The data set `noisy_circle_tda_rips` is a persistence diagram computed using the
**TDA** package with the
[`TDA::ripsDiag()`](https://www.rdocumentation.org/packages/TDA/versions/1.9.1/topics/ripsDiag)
and stored as an object of class 'diagram', which is a matrix with 3 columns:

```{r}
head(noisy_circle_tda_rips)
```

The data structure adopted by the *tdaverse* suite of packages and provided by
this package is designed to be a common interface for persistence data. This
allows for a seamless integration of persistence data across different packages
and ensures that the data is always in a consistent format.

Specifically, the 'persistence' class is a list with the following components:

- `pairs`: A list of 2-column matrices containing birth-death pairs. The
\eqn{i}-*th* element of the list corresponds to the \eqn{(i-1)}-*th* homology
dimension. If there is no pairs for a given dimension but there are pairs in
higher dimensions, the corresponding element(s) is/are filled with a
\eqn{0 \times 2} numeric matrix.
- `metadata`: A list of 3 elements containing information about how the data
was computed:

    - `data`: The name of the object containing the original data on which the
    persistence data was computed.
    - `engine`: The name of the package that computed the persistence data.
    - `simplicial_complex`: The type of simplicial complex used in the
    computation.
    - `parameters`: A list of parameters used in the computation.
    - `call`: The exact call that generated the persistence data.

The `as_persistence()` function is used to convert persistence data from
different packages into the 'persistence' class. For example, the persistence
data computed using the **ripserr** package can be converted into the
'persistence' class as follows:

```{r}
as_persistence(noisy_circle_ripserr)
```

Similarly, the persistence data computed using the **TDA** package can be
converted into the 'persistence' class as follows:

```{r}
as_persistence(noisy_circle_tda_rips)
```
