---
title: "Validation and Benchmark of Wasserstein Distances"
vignette: >
  %\VignetteIndexEntry{Validation and Benchmark of Wasserstein Distances}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{bench,ggplot2,scales,TDA}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
bibliography: references.bib
---

```{r}
#| label: setup
library(ggplot2)
library(phutil)
```

This vignette introduces the Wasserstein and bottleneck distances between
persistence diagrams and their implementations in {phutil}, adapted from
[Hera](https://github.com/anigmetov/hera), by way of two tasks:

1. Validate the implementations on an example computed by hand.
2. Benchmark the implementations against those provided by {TDA} (adapted from
Dionysus).

We will also use the {tdaunif} package to generate larger point clouds and the
{bench} package to perform benchmark tests.

## Definitions

_Persistence diagrams_ are multisets (sets with multiplicity) of points in the
plane that encode the interval decompositions of persistent modules obtained
from filtrations of data (e.g. Vietoris--Rips filtrations of point clouds and
cubical filtrations of numerical arrays).
Most applications consider only ordinary persistent homology, so that all points
live in the upper-half plane; and most involve non-negative-valued filtrations,
so that all points live in the first quadrant. The examples in this vignette
will be no exceptions.

We'll distinguish between persistence diagrams, which encode usually encode one
degree of a persistence module, and _persistence data_, which comprises
persistent pairs of many degrees (and annotated as such). Whereas a diagram is
typically represented as a 2-column matrix with columns for birth and death
values, data are typically represented as a 3-column matrix with an additional
column for (whole number) degrees.

The most common distance metrics between persistence diagrams exploit the family
of _Minkowski distances_ $D_p$ between points in $\mathbb{R}^n$ defined, for $1
\leq p < \infty$, as follows:

$$ D_p(x,y) = \left(\sum_{i=1}^{n}{(x_i - y_i)^p}\right)^{1/p}. $$

In the limit $p \to \infty$, this expression approaches the following auxiliary
definition:

$$ D_\infty(x,y) = \max_{i=1}^{n}{\lvert x_i - y_i \rvert}. $$

As the parameter $p$ ranges between $1$ and $\infty$, three of its values yield
familiar distance metrics: The taxicab distance $D_1$, the Pythagorean distance
$D_2$, and the Chebyshev distance $D_\infty$.

The [_Kantorovich_ or _Wasserstein metric_](https://en.wikipedia.org/wiki/Wasserstein_metric) derives from the problem
of optimal transport: What is the minimum cost of relocating one distribution to
another? In the space of persistence diagrams, only finitely many points lie off
the diagonal, though every point on the diagonal is taken to be included with
multiplicity $\mathbb{N}$. So the cost of relocating one diagram $X$ to another
$Y$ amounts to (a) the cost of relocating some off-diagonal points to other
off-diagonal points plus (b) the cost of relocating the remaining off-diagonal
points to the diagonal (and vice-versa). This cost then depends entirely on how
the off-diagonal points are matched, where a _matching_ is any bijective map
$\varphi : X \to Y$, though in practice we assume that almost all diagonal
points are matched to themselves and incur no cost.

The cost $D(x,\varphi(x))$ of relocating a point $x$ to its matched point
$\varphi(x)$ is typically taken to be a Minkowski distance $D_q(x,\varphi(x)) =
\lVert x - \varphi(x) \rVert_q$, defined by the $L^q$ norm on $\mathbb{R}^2$.
(While simple, this geometric treatment elides that the points in the plane
encode the collection of interval modules into which the persistence module
decomposes. Other metrics have been proposed for this space, but we restrict to
this family here.)

The total cost of the relocation is canonically taken to be the Minkowski
distance $\left( \sum_{x \in X}{D(x,\varphi(x))^p} \right)^{1/p}$ of the vector
of matched-point distances. The Wasserstein distance is defined to be the
infimum of this value over all possible matches.
This yields the formulae

$$ W_p^q(X,Y) = \inf_{\varphi : X \to Y}{\left( \sum_{x \in X}{{\lVert
x-\varphi(x) \rVert_q}^p} \right)^{1/p}}, $$

for $p < \infty$ and

$$ W_\infty^q(X,Y) = \inf_{\varphi : X \to Y}{\max_{x \in X}{\lVert x-\varphi(x)
\rVert_q}} $$

for $p = \infty$.

See @cohen2010lipschitz and @bubenik2023exact for detailed treatments and
stability results on these families of metrics.

## Validation

### Distances between nontrivial diagrams

The following persistence diagrams provide a tractable example:

$$ X = \left[ \begin{array}{cc} 1 & 3 \\ 3 & 5 \end{array} \right], \phantom{X =
Y} Y = \left[ \begin{array}{cc} 3 & 4 \end{array} \right]. $$

For convenience in the code, we omit dimensionality and focus only on the matrix
representations.

```{r define small PDs}
X <- rbind(
  c(1, 3),
  c(3, 5)
)
Y <- rbind(
  c(3, 4)
)
```

We overlay both diagrams in @fig-plot-small. Note that the vector between the
off-diagonal points $(1,3)$ of $X$ and $(3,4)$ of $Y$ is $(2,1)$, while the
vector from $(1,3)$ to its nearest diagonal point $(2,2)$ is $(1,-1)$. That one
coordinate is the same size while the other is smaller implies that an optimal
matching will always match $(1,3)$ with the diagonal, so long as $p \geq 1$. A
similar argument necessitates that $(3,4)$ of $Y$ must match with $(3,5)$ of
$X$.

```{r}
#| label: fig-plot-small
#| fig-width: 4
#| fig-height: 4
#| fig-align: center
#| fig-cap: "Overlaid persistence diagrams $X$ (circles) and $Y$ (diamond) with dashed segments connecting optimally matched pairs."
par(mar = c(4, 4, 1, 1) + .1)
plot(
  NA_real_,
  xlim = c(0, 6), ylim = c(0, 6), asp = 1, xlab = "birth", ylab = "death"
)
abline(a = 0, b = 1)
points(X, pch = 1)
points(Y, pch = 5)
segments(X[, 1], X[, 2], c(2, Y[, 1]), c(2, Y[, 2]), lty = 2)
par(mar = c(5, 4, 4, 2) + .1)
```

Based on these observations, we get this expression for the Wasserstein distance
using the $q$-norm half-plane metric and the $p$-norm "matched space" metric:

$$ W_p^q(X,Y) = ( {\lVert a \rVert_q}^p + {\lVert b \rVert_q}^p )^{1/p}, $$

where $a = (1,-1)$ and $b = (0,-1)$ are the vectors between matched points. We
can now calculate Wasserstein distances "by hand"; we'll consider those using
the half-plane Minkowski metrics with $q=1,2,\infty$ and the "matched space"
metrics with $p=1,2,\infty$.

First, with $q=1$, we get $\lVert a \rVert_q = 1+1=2$ and $\lVert b \rVert_q =
0+1=1$. So the $(1,p)$-Wasserstein distance will be the $p$-Minkowski norm of
the vector $(2,1)$, given by $W_p^1(X,Y) = (2^p + 1^p)^{1/p}$. This nets us the
values $W_1^1(X,Y) = 3$ and $W_2^1(X,Y) = \sqrt{5}$. And then $W_\infty^1(X,Y) =
\max(2,1) = 2$. The reader is invited to complete the rest of @tbl-small.

| Metric | $\lVert a \rVert$ | $\lVert b \rVert$ | $W_1$ | $W_2$ | $W_\infty$ |
|:-------|:----------------:|:----------------:|:-------:|:-------:|:------------:|
| $L^1$ | 2 | 1 | 3 | $\sqrt{5}$ | 2 |
| $L^2$ | $\sqrt{2}$ | 1 | $1+\sqrt{2}$ | $\sqrt{3}$ | $\sqrt{2}$ |
| $L^\infty$ | 1 | 1 | 2 | $\sqrt{2}$ | 1 |

: Distances between optimally paired features and Wasserstein distances between
$X$ and $Y$ for several choices of half-plane and "matched space" metrics.
{#tbl-small}

The results make intuitive sense; for example, the values change monotonically
along each row and column.
Let us now validate the bottom row---using the $L^\infty$ distance on the
half-plane, giving the popular _bottleneck distance_---using both Hela, as
exposed through {phutil}, and Dionysus, as exposed through {TDA}:

```{r validate small PDs with Hera}
wasserstein_distance(X, Y, p = 1)
wasserstein_distance(X, Y, p = 2)
bottleneck_distance(X, Y)
```

In order to compute distances with {TDA}, we must restructure the PDs to include
a `"dimension"` column.
Note also that `TDA::wasserstein()` does not take the $1/p$th power after
computing the sum of $p$th powers; we do this manually to get comparable
results:

```{r validate small PDs with Dionysus}
TDA::wasserstein(cbind(0, X), cbind(0, Y), p = 1, dimension = 0)
sqrt(TDA::wasserstein(cbind(0, X), cbind(0, Y), p = 2, dimension = 0))
TDA::bottleneck(cbind(0, X), cbind(0, Y), dimension = 0)
```

### Distances from the trivial diagram

An important edge case is when one persistence diagram is trivial, i.e. contains
only the diagonal so is "empty" of off-diagonal points.
This can occur unexpectedly in comparisons of persistence data, as the data may
be large but higher-degree features present in one set but absent in another.
To validate the distances in this case, we create an empty diagram $E$ and use
the same code to compare it to $X$. The point $(3,5)$ of $X$ will be matched to
the diagonal $(4,4)$, which yields the same $\infty$-distance $1$ so the
$L^\infty$ Wasserstein distances will be the same as before.

```{r validate small PD vs empty}
# empty PD
E <- matrix(NA_real_, nrow = 0, ncol = 2)
# with dimension column
E_ <- cbind(matrix(NA_real_, nrow = 0, ncol = 1), E)
# distance from empty using phutil/Hera
wasserstein_distance(E, X, p = 1)
wasserstein_distance(E, X, p = 2)
bottleneck_distance(E, X)
# distance from empty using TDA/Dionysus
TDA::wasserstein(E_, cbind(0, X), p = 1, dimension = 0)
sqrt(TDA::wasserstein(E_, cbind(0, X), p = 2, dimension = 0))
TDA::bottleneck(E_, cbind(0, X), dimension = 0)
```

## Benchmarks

For a straightforward benchmark test, we compute PDs from point clouds sampled
with noise from two one-dimensional manifolds embedded in $\mathbb{R}^3$: the
circle as a trefoil knot and the segment as a two-armed archimedian spiral.
To prevent the results from being sensitive to an accident of a single sample,
we generate lists of 24 samples and benchmark only one iteration of each
function on each.

```{r compute large PDs fake, eval=FALSE}
set.seed(28415)
n <- 24
PDs1 <- lapply(seq(n), function(i) {
  S1 <- tdaunif::sample_trefoil(n = 120, sd = .05)
  as_persistence(TDA::ripsDiag(S1, maxdimension = 2, maxscale = 6))
})
PDs2 <- lapply(seq(n), function(i) {
  S2 <- cbind(tdaunif::sample_arch_spiral(n = 120, arms = 2), 0)
  S2 <- tdaunif::add_noise(S2, sd = .05)
  as_persistence(TDA::ripsDiag(S2, maxdimension = 2, maxscale = 6))
})
```

```{r compute large PDs true, echo=FALSE}
n <- 24
PDs1 <- trefoils
PDs2 <- arch_spirals
```

Both implementations are used to compute distances between successive pairs of
diagrams. The computations are annotated by homological degree and Wasserstein
power so that these results can be compared separately.

```{r}
#| eval: false
# FIXME: Ensure that empty PDs (dimension 2 below) are handled correctly.
# Apparently this is an issue in TDA, not phutil
# FIXME: Replace with `seq(0, 2)` once bugs are resolved.
```

```{r benchmark phutil and TDA}
PDs1_ <- lapply(lapply(PDs1, as.data.frame), as.matrix)
PDs2_ <- lapply(lapply(PDs2, as.data.frame), as.matrix)
# iterate over homological degrees and Wasserstein powers
bm_all <- list()
PDs_i <- seq_along(PDs1)
for (dimension in seq(0, 1)) {
  # compute
  bm_1 <- do.call(rbind, lapply(seq_along(PDs1), function(i) {
    bench::mark(
      TDA = TDA::wasserstein(
        PDs1_[[i]], PDs2_[[i]], 
        dimension = dimension, 
        p = 1
      ),
      phutil = wasserstein_distance(
        PDs1[[i]], PDs2[[i]], 
        dimension = dimension, 
        p = 1
      ),
      iterations = 1
    )
  }))
  bm_2 <- do.call(rbind, lapply(seq_along(PDs1), function(i) {
    bench::mark(
      TDA = sqrt(TDA::wasserstein(
        PDs1_[[i]], PDs2_[[i]], 
        dimension = dimension, 
        p = 2
      )),
      phutil = wasserstein_distance(
        PDs1[[i]], PDs2[[i]], 
        dimension = dimension, 
        p = 2
      ),
      iterations = 1
    )
  }))
  bm_inf <- do.call(rbind, lapply(seq_along(PDs1), function(i) {
    bench::mark(
      TDA = TDA::bottleneck(
        PDs1_[[i]], PDs2_[[i]], 
        dimension = dimension
      ),
      phutil = bottleneck_distance(
        PDs1[[i]], PDs2[[i]], 
        dimension = dimension
      ),
      iterations = 1
    )
  }))
  # annotate
  bm_1$power <- 1; bm_2$power <- 2; bm_inf$power <- Inf
  # combine
  bm_res <- rbind(bm_1, bm_2, bm_inf)
  bm_res$degree <- dimension
  # collate
  bm_all <- c(bm_all, list(bm_res))
}
bm_all <- do.call(rbind, bm_all)
```

@fig-benchmark-large compares the distributions of runtimes by homological
degree (column) and Wasserstein power (row). We note that Dionysus via {TDA}
clearly outperforms Hera via {phutil} on degree-1 and -2 PDs, which in these
cases have many fewer features. However, the tables are turned in degree 0, in
which the PDs have many more features. While by no means exhaustive and not
necessarily representative, these results suggest that Hera via {phutil} scales
more efficiently than Dionysus via {TDA} and should therefore be preferred for
projects involving more feature-rich data sets.

```{r}
#| label: fig-benchmark-large
#| fig-width: 6
#| fig-height: 3
#| fig-align: 'center'
#| fig-retina: 2
#| fig-cap: "Benchmark comparison of Dionysus via {TDA} and Hera via {phutil} on
#| large persistence diagrams: Violin plots of runtime distributions on a common
#| scale."
bm_all <- transform(
  bm_all, 
  implementation = as.character(expression), 
  time = unlist(time)
)
bm_all <- subset(bm_all, select = c(implementation, degree, power, time))
ggplot(bm_all, aes(x = time, y = implementation)) +
  facet_grid(
    rows = vars(power), cols = vars(degree),
    labeller = label_both
  ) +
  geom_violin() +
  scale_x_continuous(transform = "log10", labels = scales::label_timespan()) +
  labs(x = NULL, y = NULL)
```

## References
